---
title: "Population Understanding and Foundations of Probability"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format:
  revealjs:
    theme: simple
    slide-number: true
    chalkboard: true
    transition: fade
execute:
  echo: true
  warning: false
  message: false
---


### Introduction
<div style="font-size:0.7em">
In biomedical and clinical science, knowledge and insights emerge from laboratory experiments, clinical trials, population studies, or genomic databases, the conclusions we draw are anchored in observed data

- the data often require interpretation:
  - determining whether a treatment effect is meaningful
  - whether a biomarker signal is reproducible
  - whether a genomic variant is rare or common
  - whether an observed association reflects biology rather than study design
 
- statistics provides the framework for this interpretation and allows researchers to understand, evaluate, and generate biomedical evidence in a systematic way
</div>

---

### Statistics and Scientific Literature
<div style="font-size:0.7em">
In scientific literature [i.e.: Science, Cell, Nature] readers often encounter metrics such as: 
  
    effect sizes, risk ratios, regression coefficients, confidence intervals, survival curves

- metrics that express strength of evidence between associations and uncertainty

- statistics in biomedicine provide the tools for distinguishing meaningful biological signals from background variation

</div>
---

### Statistics and Clincial Science
<div style="font-size:0.7em">

- study design decisions shape the type and strength of evidence produced
- Example study design decisions:
  - cohort selection
  - measurement strategies
  - analytical approaches
- statistical frameworks guide these decisions by clarifying how data can support specific scientific questions and how uncertainty will be quantified
- designing a clinical trial involves anticipating variability and determining sample size requirements
- designing a biomarker study involves assessing measurement reliability and effect magnitude
- statistical approaches help define these choices

</div>
---

### Statistics and Biomedical Science

<div style="font-size:0.7em">

Biomedical research often begins with a question that invovles determining:

> What does our select dataset tell us about a broader population?

Scientific task generates numbers and our goal is to interpret those numbers as evidence

Approach for analysis would involve describing:

- the scientific target of interest  
- the observed data used as evidence  
- the quantity the analysis aims to estimate  
- the uncertainty that accompanies any conclusion  

Next, we introduce the core terms that help define statistical reasoning in biomedical science

</div>

---

### Core Statistical Definitions
<div style="font-size:0.7em">
- define terms within the basic framework for statistical inference, regression modeling, and biomedical decision-making

- terms include:

  - population
  - sample
  - parameter
  - statistic

</div>
---

### Core Statistical Definitions
<div style="font-size:0.7em">
- define terms within the basic framework for statistical inference, regression modeling, and biomedical decision-making

- **population**: complete set of biological units, measurements, or possible outcomes relevant to a scientific question

- population is often entirely unobservable

**Examples**

- all patients with pancreatic cancer in the United States
- all circulating tumor cells in a single patient’s bloodstream
- all potential gene expression profiles under a defined biological condition

- statistical test exist because we cannot measure every unit in the population; instead, we draw inferences from samples

</div>
---

### Sample: Observable Subset of the Population
<div style="font-size:0.7em">

**Sample**

- subset of the population that we actually observe and measure
- sample: **data available for analysis**
- statistical conclusions originate from this subset

Statistical inference answers:

> What does this sample tell us about the population?

Examples:

- 200 patients enrolled in a clinical trial
- 50 tumor biopsies selected for sequencing
- 10,000 cells measured in single‑cell RNA‑seq

</div>
---

### Random Sampling and Statistical Inference

<div style="font-size:0.7em">

- random sampling: a data collection process in which each unit in the target population has a known, non-zero probability of selection determined by chance rather than investigator choice

Random sampling helps ensure:

  - absence of systematic selection bias
  - representation of population heterogeneity
  - quantifiable sampling uncertainty
  - validity of probability-based inference

Consequence of non-random sampling

  - estimates may be biased
  - variability cannot be quantified correctly
  - population inference may be invalid

</div>
---

### Sampling Variability and Behavior of Sample Statistics

<div style="font-size:0.7em">

**Sampling variability**: 
- inherent variation in a statistic across repeated random samples from the same population
- observable consequence of random sampling and represents the distribution of possible estimates that could have been obtained under repeated data collection

- sampling variability is linked to:

  - confidence intervals $\rightarrow$ quantifying estimation uncertainty
  - hypothesis testing $\rightarrow$ assessing compatibility with models
  - regression inference $\rightarrow$ evaluating parameter uncertainty
</div>
---

### Statistics and Variation Across Samples

<div style="font-size:0.7em">

- **sampling variability**: variation in a statistic across repeated samples drawn from the same population

- value reflects the randomness inherent in sampling rather than true differences in the underlying population

- **scenario**:  repeat a study and collect a new sample

- **outcome**: obtain a different statistic

Example:

Study 1: response rate = 42%  
Study 2: response rate = 47%  

</div>
---


### Statistical Inference Requires Sampling

<div style="font-size:0.7em">
Measuring the entire population is typically impossible due to:
- Cost
- Time
- Ethical constraints
- Biological feasibility

Examples:

- Cannot sequence every cancer cell in a tumor
- Cannot enroll every patient in a clinical trial
- Cannot observe every biological realization

- sampling makes statistical inference feasible
</div>
---

### Parameter: Linking Population to Sample

<div style="font-size:0.7em">

**Parameter**: fixed but unknown numerical characteristic of the population

**Examples**:

- True cancer prevalence rate
- True mean gene expression level
- True treatment response probability

**Nomenclature examples**:

$$
\mu, \; \sigma, \; p, \; \beta
$$
</div>
---



### Statistic: Observable Estimate of the Parameter

<div style="font-size:0.7em">

- **statistic**: computed from the sample and used to estimate the parameter

- *statistics*: random variables because they depend on the sampled data

**Examples**:

- Sample mean: $\bar{x}$ estimates population mean $\mu$
- Sample proportion: $\hat{p}$ estimates population proportion $p$
- Regression coefficients estimate population relationships

</div>
---

### Definition: Statistic
<div style="font-size:0.7em">
- **statistic**: numerical summary computed from a sample

- a statistic describes what we observed in the data

Examples:

- Sample mean age of patients: 63.4 years  
- Sample response rate: 42%  
- Observed allele frequency in gnomAD: 0.003  

- Note: these numbers summarize the sample — not the entire population
</div>
---

### Statistic vs Parameter
<div style="font-size:0.7em">

**parameter**: 

- fixed numerical characteristic of the entire population
- value does not change 
- value typically cannot observed directly

Example: true mean gene expression under a condition $\mu$

**statistic**:

- numerical summary computed from sample data
- value can vary from sample to sample
- serves as an estimate of the corresponding parameter

Example: sample mean gene expression under a condition $\bar{\mu}$

Note:

- never directly observe parameters
- observe statistics $\rightarrow$ use probability theory to understand how those statistics relate to the underlying parameters
</div>
---



### Example: Biomedical Example: Sampling Variability in Population Genomics

<div style="font-size:0.7em">

**Background: What is allele frequency?**

- **allele frequency**: proportion of all gene copies at a specific genomic locus within a defined population that carry a particular allele
- **allele frequency**: reflects how common that genetic variant is in that population

Allele frequency is therefore calculated as:


$$
\text{Allele frequency} =
\frac{\text{Number of occurrences of a given allele}}
{\text{Total number of alleles at that locus in the population}}
$$

</div>

---

### Example: Biomedical Example: Sampling Variability in Population Genomics

<div style="font-size:0.7em">



**Population parameter**

- true allele frequency of a pathogenic variant among all humans
- represents a biological property of the population
- fixed but unknown because sequencing every human is infeasible


**Statistic**

- observed allele frequency calculated from a sample dataset [i.e.:gnomAD ]
- population databases [i.e.: gnomAD] are large but finite samples
- serves as an estimate of the population parameter
- value depends on which individuals were included in the sample

</div>

---

### Example: Biomedical Example: Sampling Variability in Population Genomics

<div style="font-size:0.7em">

**Background**

- different sequencing cohorts produce slightly different estimates of allele frequency
- population databases therefore provide statistical estimates of population genetic parameters rather than exact values

**Central question**:
Why do allele frequency estimates vary across sequencing datasets?

**Approach**:
Simulate repeated sequencing studies from a population with a fixed allele frequency


</div>

---

### Example: Biomedical Example: Sampling Variability in Population Genomics

<div style="font-size:0.3em">

**Conclusions**

- observed allele frequencies fluctuate across samples, illustrating sampling variability
- figure shows: sampling variability of allele frequency estimates across repeated sequencing studies
- histogram shows : repeated sequencing studies yield slightly different allele frequency estimates that cluster around the true population frequency $\rightarrow$ sampling variability
- red line [population parameter]: actual underlying allele frequency
- even when the true allele frequency is fixed, different samples produce slightly different estimates

**Reasons for differences**:

- each sequencing cohort contains a different set of chromosomes
- random sampling causes small differences in observed variant counts

```{r, echo = FALSE}
# True population parameter (unknown in practice)
true_freq <- 0.01   # 1% pathogenic variant frequency

# Sample size similar to a population database subset
n <- 10000

# Repeat sampling to illustrate sampling variability
set.seed(123)
estimates <- replicate(50, mean(rbinom(n, size = 1, prob = true_freq)))

# Visualize sampling variability
plot <- hist(estimates,
     main = "Sampling variability of variant frequency estimates",
     xlab = "Estimated allele frequency",
     breaks = 15)

# Add true population parameter for reference
abline(v = true_freq, col = "red", lwd = 2)
```


```{r}
plot
```

</div>

---

### Conceptual Model of Statistical Inference

<div style="font-size:0.7em">
- statistical inference is a process that uses observed data to learn about unknown features of a population

- translates sample-derived summaries into probabilistic statements about population parameters while accounting for sampling variability

Workflow:

Population (unknown parameter)
$\rightarrow$
Random sampling
$\rightarrow$
Observed sample
$\rightarrow$
Compute statistic
$\rightarrow$
Estimate population parameter
$\rightarrow$
Quantify uncertainty

</div>
---

### Concept Review

<div style="font-size:0.7em">

- **population**: 
  - population contains the biological or clinical quantity of interest — such as a response probability, allele frequency, or survival rate
  - quantity is a parameter $\rightarrow$ fixed but unknown

- **random sampling**: 
  - subset of units [patients, genomes, cells] selected from the population
  - sampling introduces randomness $\rightarrow$ only some population members are observed
  
- **observed sample**: collected dataset represents one realization from many possible samples that could have been obtained

</div>
---

### Concept Review

<div style="font-size:0.7em">

- **compute statistic**:statistic [i.e.: proportion, regression coefficient, hazard ratio] summarizes information in the sample

- **estimate population parameter**: 
  - statistic serves as an estimate of the population parameter 
  - evidence-based approximation of the underlying biological quantity
  
- **quantify uncertainty**: 
  - statistic depends on the sample, it is subject to sampling variability
  - statistical inference therefore quantifies uncertainty using tools such as standard errors, confidence intervals, likelihood, or posterior distributions
</div>
---


